Machine Learning Methods for Discourse Segmentation of Communications in E-Mail Based Behavioral Interventions

### 2018-07-27 paper-amia-summit-2018/manuscript.tex
added figures for parameters

We utilized three types of features in conjunction with CRF, MLP, BRNN and CRNN methods: word embeddings, punctuation and POS features. Since POS tags have been shown to be effective semantic abstractions of individual words, we used topic features for our experiment \cite{liu2005using,treviso2017sentence}. To extract our POS features, we tag the e-Coaching emails using the NLTK POS tagger \cite{bird2009natural}. Punctuation marks, which correspond to one of the symbols \{`.', `,', `!', `?', `:', `;'\} between a pair of words, are also employed as a feature since punctuation marks designate the boundary of a sentence, clause and phrase and often also correspond to a segment boundary. For natural language processing (NLP) tasks, inputs are received as a text where words can be considered as the basic processing unit. Therefore, it is important to represent a word in such a way that it carries all relevant information. Word embeddings are one of this representation where each word represented as a real-valued vector in a high dimensional vector space. The distributed representation of word allows capturing semantic, syntactic and morphological information from large unannotated corpora  \cite{pennington2014glove, mikolov2013distributed}. In our experiment, we utilized pre-trained word2vec word vectors for words and trained word2vec word vectors for punctuation marks.

\begin{figure}[!htb]
    \centering
    \includegraphics[width=1.0\textwidth]{figures/roc-curves.eps}
    \caption{\textbf{Receiver operating characteristic curves showing the performance of binary classifiers for the segmentation of e-Coaching text when lexical features (left) and combination of lexical, DL-LDA-based topic (except LSTM and GRU) and punctuation features (right) are used}.}
    \label{fig:roc-curves22}
\end{figure}

@@

In the present study, we experimented with four different classifiers, including one traditional machine learning method, CRF and three neural networks models. Since neural networks provide a flexible architecture for constructing and synthesizing complex models, we take the advantage of this flexibility to test MLP, BRNN and CRNN models for the task of segmentation of e-Coaching emails.

\begin{figure}[!htb]
    \centering
    \includegraphics[width=1.0\textwidth]{figures/roc-curves.eps}
    \caption{\textbf{Receiver operating characteristic curves showing the performance of binary classifiers for the segmentation of e-Coaching text when lexical features (left) and combination of lexical, DL-LDA-based topic (except LSTM and GRU) and punctuation features (right) are used}.}
    \label{fig:roc-curves22}
\end{figure}

@@

\begin{figure}[!htb]
    \centering
    \includegraphics[width=1.0\textwidth]{figures/CRNN.eps}
    \caption{\textbf{Receiver operating characteristic curves showing the performance of binary classifiers for the segmentation of e-Coaching text when lexical features (left) and combination of lexical, DL-LDA-based topic (except LSTM and GRU) and punctuation features (right) are used}.}
    \label{fig:crnn}
\end{figure}

@@

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM and KNN methods for identification of ``new segment'' and ``same segment'' classes as well as their weighted average. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_base}
  \begin{tabular}{|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{new segment}} & \multicolumn{3}{|c|}{\textbf{weighted macro average}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} \\ \hline    
 CRF & 0.782 & 0.691 & 0.733 & 0.983 & 0.984 & 0.984 \\ \hline
 MLP & \textbf{0.836} & 0.593 & 0.694 & 0.982 & 0.983 & 0.982 \\ \hline
 BRNN & 0.606 & 0.680 & 0.641 & 0.977 & 0.976 & 0.976 \\ \hline
 CRNN & 0.775 & \textbf{0.797} & \textbf{0.785} & \textbf{0.986} & \textbf{0.986} & \textbf{0.986} \\ \hline
  \end{tabular}
\end{table}

@@

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM and KNN methods with different topic model-based features for segmentation of e-Coaching emails as a weighted average over ``same segment'' and ``new segment'' classification results, when all features are used together. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_weighted_avg}
 \begin{tabular}{|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{new segment}} & \multicolumn{3}{|c|}{\textbf{weighted macro average}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} \\ \hline    
 CRF & 0.813 & 0.772 & 0.792 & 0.988 & 0.988 & 0.988 \\ \hline
 MLP & \textbf{0.814} & 0.732 & 0.771 & 0.987 & 0.987 & 0.987 \\ \hline
 BRNN & 0.683 & 0.820 & 0.745 & 0.985 & 0.983 & 0.984 \\ \hline
 CRNN & 0.789 & \textbf{0.864} & \textbf{0.825} & \textbf{0.990} & \textbf{0.989} & \textbf{0.989} \\ \hline
  \end{tabular}
\end{table}

@@

Punctuation mark and topic model-based features made a significant improvement in performance of all machine learning methods. Among all topic model-based feature, the DL-LDA-based topic feature provides the highest performance in ``new segment'' and ``same segment'' classification as well as weighted average over ``new segment'' and ``same segment'' classification results. In nearly all cases, ML methods perform better, when lexical features are used in combination with punctuation and topic model-based features. This results also indicate that segmentation performance might be improved by adding additional relevant features.

\begin{figure}[!htb]
    \centering
    \includegraphics[width=1.0\textwidth]{figures/roc-curves.eps}
    \caption{\textbf{Receiver operating characteristic curves showing the performance of binary classifiers for the segmentation of e-Coaching text when lexical features (left) and combination of lexical, DL-LDA-based topic (except LSTM and GRU) and punctuation features (right) are used}.}
    \label{fig:roc-curves}
\end{figure}

@@

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM and KNN methods with different topic model-based features for segmentation of e-Coaching emails as a weighted average over ``same segment'' and ``new segment'' classification results, when all features are used together. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_weighted_avg}
 \begin{tabular}{|l|l|l|l|l|}
  \hline
\multirow{2}{*}{\textbf{Features}} & \multicolumn{4}{|c|}{\textbf{Area Under the Curve (AUC)}} \\\cline{2-5}
 & \textbf{CRF} & \textbf{MLP}  & \textbf{BRNN} & \textbf{CRNN} \\ \hline      
 word embeddings & \textbf{0.966} & 0.951 & 0.920 & 0.965 \\ \hline
 all features & \textbf{0.994} & 0.992 & 0.981 & 0.986 \\ \hline
  \end{tabular}
\end{table}

>>>>

We utilized three types of features in conjunction with CRF, MLP, BRNN and CRNN methods: word embeddings, punctuation and POS features. Since POS tags have been shown to be effective semantic abstractions of individual words, we used topic features for our experiment \cite{liu2005using,treviso2017sentence}. To extract our POS features, we tag the e-Coaching emails using the NLTK POS tagger \cite{bird2009natural}. Punctuation marks, which correspond to one of the symbols \{`.', `,', `!', `?', `:', `;'\} between a pair of words, are also employed as a feature since punctuation marks designate the boundary of a sentence, clause and phrase and often also correspond to a segment boundary. For natural language processing (NLP) tasks, inputs are received as a text where words can be considered as the basic processing unit. Therefore, it is important to represent a word in such a way that it carries all relevant information. Word embeddings are one of this representation where each word represented as a real-valued vector in a high dimensional vector space. The distributed representation of word allows capturing semantic, syntactic and morphological information from large unannotated corpora  \cite{pennington2014glove, mikolov2013distributed}. Since pre-trained word2vec word vectors provides the best results, in our experiment, we utilized pre-trained word2vec word vectors for words. Trained word vectors were also utilized when punctuation marks are not available in the pre-trained word2vec word vectors. Figure~\ref{fig:embedding-dimension} illustrates the performance of CRNN model on e-Coaching email segmentation by varying the dimension of pre-trained and trained word embeddings with GloVe and word2vec models.

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.6\textwidth]{figures/embedding-dimension.eps}
    \caption{\textbf{Performance of CRNN model on e-Coaching email segmentation by varying the dimension of pre-trained and trained word embeddings with GloVe and word2vec models.}}
    \label{fig:embedding-dimension}
\end{figure}

@@

In the present study, we experimented with four different classifiers, including one traditional machine learning method, CRF and three neural networks models. Since neural networks provide a flexible architecture for constructing and synthesizing complex models, we take the advantage of this flexibility to test MLP, BRNN and CRNN models for the task of segmentation of e-Coaching emails.\\

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.60\textwidth]{figures/length-mlp.eps}
    \caption{\textbf{Performance of MLP model on e-Coaching email segmentation by varying the length of sample size.}}
    \label{fig:length-mlp}
\end{figure}

@@

\begin{figure}[!htb]
    \centering
    \includegraphics[width=1.0\textwidth]{figures/CRNN.eps}
    \caption{\textbf{Architecture of convolutional recurrent neural networks for automatic segmentation of e-Coaching emails}.}
    \label{fig:crnn}
\end{figure}

@@

\begin{table}[ht]
\centering
\caption{Performance of CRF, MLP, BRNN and CRNN methods for identification of ``new segment'' class as well as weighted average over ``new segment'' and ``same segment'' classes when word embeddings or lexical features are used. The highest value for each performance metric
is highlighted in bold.}
\label{tab:result_base}
  \begin{tabular}{|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{new segment}} & \multicolumn{3}{|c|}{\textbf{weighted macro average}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} \\ \hline    
 CRF & 0.782 & 0.691 & 0.733 & 0.983 & 0.984 & 0.984 \\ \hline
 MLP & \textbf{0.836} & 0.593 & 0.694 & 0.982 & 0.983 & 0.982 \\ \hline
 BRNN & 0.606 & 0.680 & 0.641 & 0.977 & 0.976 & 0.976 \\ \hline
 CRNN & 0.775 & \textbf{0.797} & \textbf{0.785} & \textbf{0.986} & \textbf{0.986} & \textbf{0.986} \\ \hline
  \end{tabular}
\end{table}

@@

\begin{table}[ht]
\centering
\caption{Performance of CRF, MLP, BRNN and CRNN methods for identification of ``new segment'' class as well as weighted average over ``new segment'' and ``same segment'' classes when all features are used together. The highest value for each performance metric
is highlighted in bold.}
\label{tab:result_weighted_avg}
 \begin{tabular}{|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{new segment}} & \multicolumn{3}{|c|}{\textbf{weighted macro average}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} \\ \hline    
 CRF & 0.813 & 0.772 & 0.792 & 0.988 & 0.988 & 0.988 \\ \hline
 MLP & \textbf{0.814} & 0.732 & 0.771 & 0.987 & 0.987 & 0.987 \\ \hline
 BRNN & 0.683 & 0.820 & 0.745 & 0.985 & 0.983 & 0.984 \\ \hline
 CRNN & 0.789 & \textbf{0.864} & \textbf{0.825} & \textbf{0.990} & \textbf{0.989} & \textbf{0.989} \\ \hline
  \end{tabular}
\end{table}

@@

Punctuation mark and topic model-based features made a significant improvement in performance of all machine learning methods. Among all topic model-based feature, the DL-LDA-based topic feature provides the highest performance in ``new segment'' and ``same segment'' classification as well as weighted average over ``new segment'' and ``same segment'' classification results. In nearly all cases, ML methods perform better, when lexical features are used in combination with punctuation and topic model-based features. This results also indicate that segmentation performance might be improved by adding additional relevant features. \\

\begin{figure}[!htb]
    \centering
    \includegraphics[width=1.0\textwidth]{figures/roc-curves.eps}
    \caption{\textbf{Receiver operating characteristic curves showing the performance of binary classifiers for the segmentation of e-Coaching text when only word embeddings/lexical (left) and combination of word embeddings/lexical, POS and punctuation features (right) are used}.}
    \label{fig:roc-curves}
\end{figure}

@@

\begin{table}[ht]
\centering
\caption{AUC values of all classifiers when only word embeddings/lexical and combination of word embeddings/lexical, POS and punctuation features are used. Highest value for each experiment across all models is highlighted in boldface.}
\label{tab:result_roc}
 \begin{tabular}{|l|l|l|l|l|}
  \hline
\multirow{2}{*}{\textbf{Features}} & \multicolumn{4}{|c|}{\textbf{Area Under the ROC Curve (AUC)}} \\\cline{2-5}
 & \textbf{CRF} & \textbf{MLP}  & \textbf{BRNN} & \textbf{CRNN} \\ \hline      
 word embeddings or lexical features & \textbf{0.966} & 0.951 & 0.920 & 0.965 \\ \hline
 all features & \textbf{0.994} & 0.992 & 0.981 & 0.986 \\ \hline
  \end{tabular}
\end{table}
